Project Manager's Progress Report

### 2015-04-11 pnsac/newsletter/article_Feature1.tex
Prep for April 2015 issue.

\title{My Days at the Home of the Merlin Engine}
\subtitle{Part 4}
\author{Richard Lodge}

@@

One of the most mysterious departments at Rolls-Royce was the
Technical Publications Department. Word processing was unheard of in
the 1960s and every time an engine modification was made, it resulted
in much work documenting even a small change and distributing the
updated documentation to all engine users around the world. I never
spent any time working with the people in this department even though
there was a complete building devoted to the work. Compared with the
resources allocated to engine development, the technical publication
side of the business was very small and it always remained a mystery
to me.

Since retiring from my career as an accountant in Canada, I have
become active on the actual restoration work of the North
Star. Understandably, I have a special interest in working on the
Merlin engines. Nobody now working on the restoration has ever worked
on a Merlin when it was in service. It is therefore necessary at all
times to refer to the various publications and manuals we have managed
to accumulate for the North Star.

It is only now, actually working on the engine and referring
frequently to the exploded diagrams that I realize how complicated the
work of the Technical Publications Department must have been. The men
and women who worked there were largely unsung heroes in that it was
very tedious and unglamorous work which was of great importance for
the safe and trouble-free running of engines in service. I can
remember that I would hear from time to time of the difficulties of
ensuring that publication updates were actually delivered to many
remote places in the world and the considerable logistical
complications of ensuring that the packages arrived safely and
furthermore that the updates were actually inserted in the relevant
manuals, particularly as many of the engine users did not speak
English. I cannot remember now if any of the documentation was
translated into other languages although I have a recollection that we
employed people who could translate into French and Spanish.

After a couple of years I was given a new job in the Long Range
Planning Department. I was one of three accountants working in this
office under a very senior accountant. Here we looked at the costings
of the development work, the projected sales  and production costs on
new engine designs. This was the time when the Spey was going into
service and the the company had signed a fixed price contract for the
RB211 engines for the new Lockheed 1011. This was a major project,
championed by David Huddie the Chief Engineer. Computer spreadsheets
did not exist at that time - pencil and columnar paper were the order
of the day, together with efficient erasers. Long range forecasts of
25 years required many yearly columns and each time we produced
figures that the engineers did not like, we were told to rework the
numbers until we produced an acceptable answer. We were not limited in
the number of erasers we could use. The usual way of cutting costs was
to reduce the forecast development and testing time. Of course this
did not work in real life.

We were often dealing with the directors of R-R who lived on the top
floor of the Nightingale Road offices. Many of them were engineers and
they had little time for young bean counters. It was still a time of
optimism for the company. The huge contract for the Lockheed engines
had been signed and the company was still enjoying its reputation as
the company whose Merlin engines had saved Britain in the air
war. This was in 1965/6 and it was several years before our
projections of doom resulted in the bankruptcy of the
company. Fortunately I left the company in 1966 and was not involved
in what must have been a horrible process as a symbol of British
engineering excellence slowly went to the wall.

After the chaos caused by the demise of Rolls-Royce Limited, the
British government formed a new nationalized company to take over the
remains of the old one. This has been very much a success story and
the newly privatized Rolls-Royce is now a highly profitable and
innovative company. I look back on my days at R-R with nostalgia but I
now realize that much of the time we were all living in a fool's
paradise.

>>>>

\title{Operation Hawk}
%\subtitle{Part 4}
\author{Hugh A. Halladay, Legion Magazine}

\textit{Photos courtesy of Library Archives Canada and 426 Thunderbird
  Sqn, RCAF.}

@@

\begin{figure}[htbp]
   \vspace{2em}
   \centering
   %name of the graphic, without the path AND in EPS format:
   \includegraphics[scale=0.5]{jim_riddoch.eps}
   %caption of the figure 
   \caption*{\small \em Jim Riddoch.}
   %label of the figure, which has to correspond to \ref{}:
   \label{fig:jim_riddoch.eps}
\end{figure}

In 1947, while Prime Minister Mackenzie King was in London, his
Minister of Foreign Affairs (and chosen suc-cessor), Louis St-Laurent,
consented to Canadian participation on the United Nations Temporary
Commission of Korea (UNTOK). On his return to Canada, King furiously
confronted St-Laurent. He wanted no part---even at the fringes—of
Korean affairs, which he considered dangerous.

St-Laurent, in turn, threatened to resign. Ultimately, the old prime
minister backed down. The incident marked a phase in the changing of
the guard in Ottawa, and a movement towards increasing Canadian
involvement in international affairs.

Mackenzie King died on July 22, 1950. As he lay in state in the
Parliament Buildings, six North Star aircraft of No. 426 Squadron flew
over, saluting his passing as they were en route to Tacoma,
Washington, to commence a Royal Canadian Air Force transoceanic
airlift in sup-port of United Nations forces fighting in Korea.

It is not known if the former prime minister was spin-ning in his
coffin.

Although the RCAF had considerable air transport experience during
the Second World War, only the work of No. 168 (Heavy Transport)
Sqdn., delivering service mail to Europe, constituted a sustained
transatlantic operation. With the retirement of its Fortress and
Liberator aircraft, which were converted from bombers to freighters,
the RCAF briefly lost its long-range transport capability. This gap
was soon filled by the Canadair North Star, built in Montreal for the
RCAF, TCA (Trans-Canada Airlines and later Air Canada) and other civil
carriers.

The North Star was basically a Douglas DC-4, but with Rolls-Royce
Merlin in-line engines instead of Pratt-andWhitney radials. It was a
dependable aircraft, but generations of passengers and crew would
remember the tremendous noise and vibration in flight. The full story
of this is related by Larry Milberry in his 1982 book The Canadair
North Star.

Deliveries of the new aircraft to the RCAF were delayed when
priority was given to TCA to help it establish a Canadian presence in
transatlantic commercial flying. Crews of No. 426 (Thunderbird)
Sqdn. finally began North Star training early in 1948, and actual
operations on the type commenced that summer.

While the RCAF did not participate in the July 1948 to May 1949 Berlin
Airlift, a few individual aircrew did while on exchange duties with
the Royal Air Force and United States Air Force. As it happened, the
airlift seriously degraded the USAF Military Air Transport Service
(MATS) after flour and coal dust penetrated the seams of every
participating aircraft. When it was next marshalled to meet the Korean
crisis, MATS needed all the help it could get—and that was forthcoming
when the Canadian government committed No. 426 Sqdn.—its only
long-range transport unit—to the Korean campaign.

In recounting the RCAF’s transpacific operational story, one must pay
tribute to Laurence Motiuk, author of two monumental volumes, namely
Thunderbirds at War: Diary of a Bomber Squadron, and Thunderbirds for
Peace: Diary of a Transport Squadron, which set out the history of
No. 426 Sqdn. The latter work was particularly difficult to write,
given that in the 1950s, RCAF units often submitted incomplete
historical narratives and in some instances neglected to prepare any
at all.

Motiuk’s account of the postwar squadron was drawn from numerous
unofficial records, including log books as well as the sketchy unit
reports.

Almost from the moment that North Korea launched its invasion of South
Korea on June 25, 1950, No. 426 Sqdn. personnel were preparing for
overseas operations. If Canada was going to be involved, that unit was
certain to be mobilized. Cabinet approval came through on July 19,
1950. The squadron cancelled all further transport work with the
exception of vital northern resupply.

Deployment to McChord Air Force Base, Tacoma, Washington, began on the
25th. Six North Stars, 12 crews and 185 ground personnel were
dispatched. Twothirds of the aircrew were veterans of the Second
World War; the remainder were men who had joined the RCAF after
1947. Three of the aircraft left Tacoma on the 27th, arriving at
Haneda in Tokyo two days later. Operation Hawk was under way.

It might be noted that No. 426 Sqdn. was not alone in supporting
operations in Korea. Historian Carl Mills reminds us that Canadian
Pacific Airlines provided charter services to MATS from August 1950 to
March 1955, conducting 703 transpacific flights. Unlike the RCAF
operation, these involved passengers only. They lost one aircraft
(CF-CPC), a DC-4 with seven crew members and 31 passengers on July 2,
1951. The aircraft, outbound from Vancouver, disappeared near
Yakutat, Alaska; no trace was ever found.

The North Stars followed a challenging route. From McChord to
Elmendorf Air Force Base in Alaska was 1,490 miles. This was followed
by a 1,537-mile run to Shemya Air Force Base in the Aleutian
Islands. Next came the 2,104mile flight to Haneda in Tokyo, with the
aircraft making landfall at Matsushima (Honshu, Japan).

The Shemya-Matsushima leg was particularly challenging; care had to
be taken not to drift into Russian airspace, a task complicated by
Russian efforts to jam radio and navigational aids. Depending on the
season, the aircraft might return by the same route, or take a more
southerly central Pacific track from Haneda to Wake Island to Hickam
Air Force Base in Honolulu. From there the aircraft would journey to
Travis Air Force Base in San Francisco, and then on to McChord.

Initially it was hoped the unit could dispatch one sortie per day, but
this proved beyond the resources of a six-plane outfit. Eventually,
No. 426 found it could send off five aircraft per week. Not
surprisingly, its peak activity coincided with the buildup of United
Nations forces in Korea and their maintenance during the period of
mobile warfare.

Between July 28 and Dec. 31, 1950, the squadron dispatched 123
aircraft across the Pacific. In the whole of 1951 it was 193 missions,
but in 1952 it dropped to 133. Not only were the Thunderbirds
returning to more traditional duties in Canada, they were also
assuming the job of providing transport, including personnel
rotation, to Canada’s growing land and air presence in Europe. In
1953—the last full year of Hawk—it was 85 missions.

The final Hawk mission departed McChord on May 31, 1954, and ended at
the squadron’s home base at Dorval on June 9, 1954. As it wound up, an
RCAF dispatch reported that No. 426 Sqdn. had flown 599 round
trips—four of them directly to Korea itself—logged 34,000 flying
hours, carried 13,000 personnel, and airlifted 3,500 tons of
freight. The report erred on two points. Six flights (not four) had
gone directly to Korea. The other discrepancy was that there were
only 584 missions; for some unknown reason the numbered listing of
flights had jumped from “439” on Nov. 26, 1952, to “455” on Nov. 28,
1952; the inaccurate 1954 press release continues to be quoted, ad
nauseam, to this day.

Operation Hawk was accomplished with no fatalities but some close
calls and more than a few incidents. On Sept. 15, 1950, Wing Commander
C.H. Mussells, Commanding Officer of No. 426 Sqdn., departed McChord
for Elmendorf with American soldiers and anti-tank ammunition. Both
were urgently needed in Korea, so the aircraft had been permitted an
overload clearance.

Three hours after takeoff, one of the Merlin engines began overheating
through a coolant leak. Mussells shut down the engine, feathered the
propeller, and prepared to return to McChord. Then a second engine on
the same side overheated. It was also shut down. There was now no
question about getting to McChord; Mussells prepared to make an
emergency landing at Sandspit (Moresby Island).

To lighten the aircraft, Mussells dumped excess fuel. Some passengers
mistook the misty gasoline trail for smoke, panicked and crowded to
the back of the fuselage, as if that would have offered safety. This
complicated the process of trimming the aircraft for landing. A
crewman ordered everyone back to their seats, but the soldiers could
not have been reassured as the North Star flew over a crashed USAF
aircraft, its tail still sticking out of the water, during the final
approach into Sandspit. Happily, Mussells landed safely.

On April 19, 1951, a North Star piloted by Flight Lieutenant J.A. Watt
was on a route-training flight to an unfamiliar field—Ashiya—located
between Osaka and Kobe, Japan. He was cleared by Ashiya tower to
descend from 4,000 to 3,000 feet. At 3,400 feet the aircraft hit trees
atop a hill that was not even indicated on the airfield map. No one
was injured but the aircraft sustained considerable damage to its
nose, oil cooler and exterior radio aerials. The pitot head was also
wiped out, and with it the airspeed indicator. The aircraft also lost
the use of an engine.

Fortunately, the passengers included three of the most experienced
North Star captains in the RCAF: Wing Cmdr. Mussells, Wing
Cmdr. J.K. MacDonald, and Flying Officer Robert Edwards. After
assessing the damage, Mussells ordered MacDonald and Edwards to com-
plete the landing, which they did.

On the night of Dec. 27, 1953, Squadron Leader E.L. Hare was homeward
bound from Japan in North Star 17516, bound for Shemya. The weather
was bad and the ceiling minimal. Visibility was a half mile with a
savage crosswind blowing at a 90-degree angle. The runway surface was
covered with wet snow; braking conditions were poor. On the third
landing attempt, everything came together in the worst possible
way. The North Star was blown off the runway into a gully. Everyone
survived, but 17516 was a total “write-off”–the only RCAF aircraft
lost in the course of Operation Hawk.

Canadian military policy with respect to honours and awards was very
restrictive from 1946 onwards, but rules were relaxed during the
Korean War. Personnel involved in Hawk were awarded one Officer of the
Order of the British Empire (OBE), one Member of the Order of the
British Empire (MBE), four Air Force Crosses, two Air Force Medals and
13 Queen’s Commendations. Most of these were to aircrew personnel, but
Sqdn. Ldr. W.H. Lord was granted an MBE for his work in establishing
and supervising the maintenance detachments that serviced the North
Stars wherever they alighted along the 11,000 mile route.

### 2015-04-11 pnsac/newsletter/article_ProgReport.tex
Prep for April 2015 issue.

\title{Project Manager's Progress Report}
\subtitle{December 2013}
\author{Bruce Gemmill}

@@

\section{Nr 3 Engine}
\label{sec:engines_3}

Engine 3 is nearing completion.  The supercharger and intercooler
preheat assemblies have been installed, along with the propeller
reduction gearbox.  While this work was being carried out by the
engine crew, the remainder of the volunteer work force busied
themselves completing the numerous cowl panels, pipes and hoses and
other pieces needed to complete the engine.  Two items that seemed to
take forever were the large steel exhaust shrouds.  A lot of hammering
was needed to pound out years of dents, then numerous rivets and all
the Dzus fasteners and springs needed to be replaced.  The complex
shape of the exhaust shroud made this challenging work.  Further
delaying completion was the search for suitable clear coat to protect
the steel from corrosion.  The old clearcoat we used successfully on
engines 1 and 2 is no longer available.  Mike Irvin finally located a
suitable replacement, and the shrouds were clear coated and serial
numbers were stencilled on.

\begin{figure}[htbp]
   \vspace{2em}
   \centering
   %name of the graphic, without the path AND in EPS format:
   \includegraphics[scale=0.5]{Eng3Intercooler.eps}
   %caption of the figure 
   \caption*{\small \em Garry installing the intercooler on engine 3.}
   %label of the figure, which has to correspond to \ref{}:
   \label{fig:Eng3Intercooler}
\end{figure}

The auxiliary gearbox was disassembled, and significant corrosion was
found on the bearings inside.  Bearings had to be ordered from
England.  Meanwhile, the gearbox was cleaned and painted.  At the same
time, the large DC generator was dismantled, repainted and
reassembled, as well as the air pump and the tachometer generator and
propeller synchronizer.  Once the new bearings were installed, the
gearbox was completed and the generator and air pump were attached.
Finally, the gearbox was installed on the firewall, which is now
finished, awaiting the installation of engine \#3.

\section{Crew Lounge, Galley,  and Forward Washroom}
\label{crewlounge}

Work has slowed on the forward section of the aircraft.  The repair
needed under the floor of the crew lounge is on hold until a suitable
piece of extruded aluminum can be found.  Until this is complete, the
new cushion flooring cannot be installed. This has also delayed the
installation of all of the crew lounge accessories, such as the table,
seats and secure storage bin, all of which are complete but waiting in
storage.

\section{Fuselage and Empennage}
\label{empenage}

The forward and rear belly compartments are now complete.  All the
frame pieces and hardware needed for the troop seats have been made,
and work is progressing on the seats and backs.  One section has
undergone a trial fit, and all pieces seem to fit perfectly.

\section{Planned Restoration Work 2013/14}
\label{sec:plannedwork}

The museum has advised all North Star volunteers that our work must be
put on hold for several months while a major exhibit for the 100th
anniversary of the beginning of the First World War is prepared.  Our
work space in the restoration hangar is needed to assemble and prepare
several aircraft that will be added to the display space in March.
Until then our work on the aircraft is suspended, but we don't expect
this delay to have a serious impact on the overall project.  We look
forward to resuming work in the Spring.

>>>>

\title{Project Manager's Progress Report}
\subtitle{April 2015}
\author{Bruce Gemmill}

@@

Since our last Project Report, our project has gone through much
upheaval, and work has been delayed for a number of reasons.  However,
I can report now that we are back on track and making steady progress
on several parts of the North Star. This report tracks progress since
April 2014.

\section{Restoration Shop Shutdown}
\label{sec:restoration_shop}

Engine \#3 was in the final stages of assembly when we had to shut down
our project at the end of December 2013, to make way for other museum
work, in preparation for events to mark the 100th anniversary of the
beginning of the First World War in 2014.  In particular, new exhibits
in the museum required all of the available floor space in the shop,
as well as full time work by all museum conservation staff, meaning
the volunteer crew could not be supervised as usual.

However, two volunteers, myself and Peter Trobridge, were allowed to
work on the aircraft during the shutdown between January and July
2014.  The focus was on completing the assembly of the cockpit and
crew lounge.  More on this later.

\section{Engine Frame}
\label{sec:engine_frame}

Nr 3 engine frame was completed after the installation of the engine
and supercharger.  Fitting the many cowl panels proved difficult, due
to the extensive rework that many of these items went through, causing
some alignment difficulties when fitting these panels onto the engine
frame.  This work was done after work resumed in July 2014, and the
completed engine was successfully installed on the aircraft last
summer.

%\begin{figure}[htbp]
%   \vspace{2em}
%   \centering
%   %name of the graphic, without the path AND in EPS format:
%   \includegraphics[scale=0.5]{fwd_cockpit_complete.eps}
%   %caption of the figure 
%   \caption*{\small \em Forward cockpit -- complete.}
%   %label of the figure, which has to correspond to \ref{}:
%   \label{fig:fwd_cockpit_complete}
%\end{figure}

\section{Nr 4 Engine}
\label{sec:engine_4}

The last engine to undergo restoration was removed in July and work
immediately began in the engine shop to disassemble the engine frame
and move the engine to the engine stand for restoration.  All cowl
panels, pipes, hoses and accessories have been removed and stored
until space is available to restore these items.  The priority has
been to disassemble the engine, which was completed in October, and
restoration and reassembly of engine 4 is now well underway.  The
refurbished crankshaft has been installed in the engine block, and the
connecting rods and pistons cleaned and polished.  Work has now begun
on the more complex cylinder heads and valve train assemblies.

Normally, work would also be underway to restore the engine frame, but
due to a lack of space, this will be delayed until the engine is back
together and can be moved out of the engine shop.  Some cleaning and
paint stripping has been done, but further work on the frame will
likely have to wait until more space is available for full restoration
of this large item.  This work will likely delay the completion of Nr
4 engine until 2016.

\section{Cocpit, Crew Lounge, Galley,  and Forward Washroom}
\label{crewlounge}

Since our last report, the crew lounge and galley equipment was fully
restored.  A new wall was installed in the galley, as the old wall
contained asbestos and needed to be removed for safety reasons.  The
galley cupboards were badly corroded, so a new set was built, with
only the original doors remaining.  This was installed in the galley,
along with the auxiliary hydraulic reservoir, after two sections of
the floor were patched to repair corrosion damage.  New flooring was
cut and glued in the crew lounge and washroom, then the washroom
equipment was installed, including the toilet and washroom door, and
the vanity sink and mirror.

The five crew seats that had been removed early in the project were
finally installed in the cockpit, after installing the radio operator
and navigator tables, and the appropriate seat rails.  We also
installed new cockpit curtains, and completed these crew positions by
installing a donated headset and several microphones, along with a
Morse code key at the radio position.

The seat rails and seats for the crew lounge were installed, along
with the crew table and secure stowage case.  The crew bed was
installed above the crew seats, and then the curtain rod with a new
set of curtains was installed to complete the crew lounge.  The access
door to the crew area has not been fully restored, but it was
re-installed with a new lock to secure the forward portion of the
aircraft.  The cockpit, crew lounge and forward washroom are now
complete.  This is a significant milestone for the project, and has
been very popular with select museum visitors who have been allowed
access.

\section{Fuselage and Main Cabin}
\label{sec:main_cabin}

The baggage compartments are now fully restored, and the refurbished
battery elevators were installed just aft of the nose wheel.  Work has
now begun on the main cabin.  This started with removal of various
fittings along the starboard side of the cabin, then removing the wood
paneling below the windows, and then removing the wood cargo floor
panels to expose the metal sub floor.  Most of the side panels and
some floor panels are badly damaged and will need to be replaced.  The
fittings for patient litters and troop seats were also removed.
Finally, the starboard windows were removed.  Once all the fittings
were removed, work began on repairing corrosion damage along the
exposed ribs and sub floor.

The rear washroom was also dismantled, after removing the toilet and
fittings.  The floor under the toilet was badly corroded, and the old
floor panel was removed.  This required drilling out approximately
1000 rivets!  A new panel has been made and will be installed shortly.
Portions of the corroded rear bulkhead were also removed, so that new
panels can be installed.  Some items were also removed from the tail
section, to allow for thorough cleaning and spot priming bare metal
where necessary.  During this work damage was discovered on the torque
tube that controls the movement of the rudder.  This was unexpected,
and will require removal of the tail cone to allow for a new torque
tube to be fitted.

Several badly corroded sections of the starboard side of the cabin
have been repaired, and work continues with corrosion repairs and
other cleanup while the aircraft is inside.  Part of the forward cargo
door has been removed for restoration.  The paratroop door insert is
complete, and the remaining portion of the forward cargo door is being
stripped and repaired prior to painting.  It is likely this will have
to be re-installed on the aircraft before all painting and restoration
is complete, so as not to delay it’s move outside.

A major effort was launched to restore the large heater ducts removed
from the main cabin.  These were covered with formed mica insulation
sections, which were removed earlier.  Many of these sections fell
apart and could not be reused.  After many repairs to the metal
ductwork, new sections of mica were formed in a jig and glued to the
ducts.  These duct sections are in storage waiting for completion of
the main cabin so they can be installed on the ceiling.

Work has now begun on the port windows and wall sections.  We hope to
complete the windows soon, as these must be installed before the
aircraft can be moved outside.

\section{Cabin Liners and Troop Seats}
\label{sec:liners_troop_seats}

The original cabin liners were removed from the aircraft earlier in
the project.  These had been washed and stored until repair work could
be done.  All liners were unrolled, photographed and assessed for
damage.  Some liners will be repaired, while others, particularly the
cargo door liners, need to be replaced because the fabric is too badly
deteriorated.  A sewing station has been set up in the storage hangar,
and several of the new liners have been completed.  These will need to
have new fasteners installed, then painted to match the original
liners.

The North Star was originally fitted with troop seats down both sides
of the main cabin.  These could be stowed against the wall to make
room for cargo, or unrolled and set up to carry military personnel.
The original seats were removed by the RCAF, so a decision was made to
make a new set, using funds donated by Project North Star members.
Seat fabric, steel and aluminum tubing was ordered, and the troop seat
sections were fabricated, based on available drawings.  All seat
sections have been completed, and have been stored until the main
cabin is completed.

\section{Planned Restoration Work---2015}
\label{sec:plannedwork}

Over the next year, we will continue to work on Engine 4 and the main
cabin and tail section.  The aircraft will be moved outside this
summer for some interior painting.  This means the windows and doors
will need to be installed for weatherproofing and security.  During
this time we also hope to begin fabricating new wall and floor
sections.  One other work item that may be started is to clean and
restore the four engine nacelles.  Pipes and fittings need to be
removed, followed by a thorough cleaning and a fresh coat of paint.
This work may be delayed until qualified volunteers become available.
